Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-07 paper/paper2.tex
more fixes to reviewer-comments
More citations and small tweaks

\section{Background}
\subsection{Markov State Models}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

Introduction to MSMs via the propagator theory motivates the view of the parameterization problem as one of \emph{function estimation}. This perspective has a critical disadvantage: the natural error metric -- the difference between the estimated MSM and true propagator is intrinsically uncomputable. The only access to the true propagator is via samples from the Hamiltonian with MD. Therefore, a compelling alternative is to view the problem as one of \emph{prediction}. Since the MSM is a generative model for trajectories in $\Omega$, it is possible to sample ``pseudo-trajectories" directly from the MSM. If the MSM were perfect, these ``pseudo-trajectories" would be indistinguishable from trajectories generated by the true propagator. Thus, the probability assigned by the MSM, to samples from the true propagator is a measure of the MSM's accuracy. This argument is formalized by Bayes' rule, which establishes a proportionality between the probability of a model given data to the probability of the data given the model
\begin{equation}
\label{eq:bayes}
P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big) = P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big) \frac{P(\{S, \mathbf{T}\})}{P(D)},
\end{equation}
where $P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big)$ is the posterior probability of an MSM, $P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big)$ is the \emph{likelihood} of the data given a model, $P(\{S, \mathbf{T}\})$ is the prior probability of an MSM before observing any data, and $P(D)$ is the probability of the data, a constant.

\subsection{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation} 
where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set. The AIC and BIC are augmented likelihood function approaches that do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
\label{eq:aic}
\mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed. Model selection is performed by a minimization of the criterion.

%\section{Prior Work}

%Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
%$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
%$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related. Specifically, when building multiple models for the same data set, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
%$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data. It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
%$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM. Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Brownian dynamics on the two dimensional M\"{u}ller potential \cite{Muller1979Location}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\label{fig:pics}
\end{figure}

@@

For all models, we computed the BIC and AIC according to \cref{eq:bic} and \cref{eq:aid}. For the BIC, we need to define the ``number of data points" in the dataset. We used the total number of transitions observed in the data as the value for $N$. However, as transitions were counted using the ``sliding window" approach, these transitions are not necessarily independent. Therefore, the BIC may be over-penalizing the number of parameters due to this over-estimate of $N$. For a discussion of the sliding window approach, we refer the reader to \citet{Prinz2011Markov}.

>>>>

\section{Theory}
\subsection{Markov State Models}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

Introduction to MSMs via the propagator theory motivates the view of the parametrization problem as one of \emph{function estimation}. This perspective has a critical disadvantage: the natural error metric -- the difference between the estimated MSM and true propagator is intrinsically uncomputable. The only access to the true propagator is via samples from the Hamiltonian with MD. Therefore, a compelling alternative is to view the problem as one of \emph{prediction}. Since the MSM is a generative model for trajectories in $\Omega$, it is possible to sample ``pseudo-trajectories" directly from the MSM. If the MSM were perfect, these ``pseudo-trajectories" would be indistinguishable from trajectories generated by the true propagator. Thus, the probability assigned by the MSM, to samples from the true propagator is a measure of the MSM's accuracy. This argument is formalized by Bayes' rule, which establishes a proportionality between the probability of a model given data to the probability of the data given the model
\begin{equation}
\label{eq:bayes}
P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big) = P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big) \frac{P(\{S, \mathbf{T}\})}{P(D)},
\end{equation}
where $P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big)$ is the posterior probability of an MSM, $P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big)$ is the \emph{likelihood} of the data given a model, $P(\{S, \mathbf{T}\})$ is the prior probability of an MSM before observing any data, and $P(D)$ is the probability of the data, a constant.

\subsection{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
\label{eq:exclusive}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
An alternative approach involving fuzzy clustering, where conformations are not uniquely assigned to a single state, would not impose the constraint in \cref{eq:exclusive}, and would also involve a likelihood function that marginalizes over all possible paths\cite{gordon1992fuzzy}. In this work, we consider only models based on crisp partitioning. As such, given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set.
\begin{equation}
  \label{eq:bic}
  \mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
  \label{eq:aic}
  \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.  Model selection is performed by a minimization of the criterion. The BIC is derived based on the dominant terms in the Laplace approximation to the logarithim of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}. The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series. Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.

%\section{Prior Work}
%Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
%$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
%$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related. Specifically, when building multiple models for the same data set, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
%$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data. It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
%$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM. Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Brownian dynamics on the two-dimensional M\"{u}ller potential \cite{Muller1979Location}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\label{fig:pics}
\end{figure}

@@

The BIC and AIC were calculated according to \cref{eq:bic} and \cref{eq:aic}, with $N$ equal to the total number of transitions observed in the data. However, as transitions were counted using the ``sliding window" approach, these transitions are not statistically independent\cite{Prinz2011Markov}. Therefore, the BIC may be over-penalizing the number of parameters due to this over-estimate of $N$.
